Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-13 paper2.qmd
section:theoretical; process:write,figures

<!-- 9. what else? -->
Building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24], the conceptual-population model then includes two hierarchical structural components. Each component defines how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010]. This addition facilitates analyses that integrate the trait measurement and hypothesis testing processes, simultaneously accounting for the traits and their associated uncertainties, thereby reducing bias and enhancing the measurement precision.

@@

@fig-cj09 displays the two hierarchical structural components in the motivating example. Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component. This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $X_{IA}$ causally influence the texts' written-quality trait $T_{IA}$. The DAG also highlights the idiosyncratic errors $e_{IA}$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain. Here, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively. Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $p_{IA}$ columns.

@@

A cautious reader will notice that these descriptions rely on several relevant variables. In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp. 37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].

@@

>>>>

<!-- 9. what else? -->
Building on the arguments developed in @sec-theory-issue2 and the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24], the conceptual-population model includes two hierarchical structural components. Each component defines how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010]. This addition facilitates analyses that integrate the trait measurement and hypothesis testing processes, simultaneously accounting for the traits and their associated uncertainties, thereby reducing bias and enhancing the measurement precision.

@@

@fig-cj09 displays the two hierarchical structural components for the motivating example. Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component. This component describes how the students' writing-quality trait $T_{I}$ and a set of *relevant* text-related variables $X_{IA}$ causally influence the texts' written-quality trait $T_{IA}$. The DAG also highlights the idiosyncratic errors $e_{IA}$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain. Here, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively. Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $p_{IA}$ columns.

@@

@@

A cautious reader will notice that the model descriptions of @sec-theory-theoretical_P rely on several relevant variables. In this study, a *relevant* or a *sufficient adjustment* set of variables (potentially an empty set) refers to those that satisfy the *backdoor criterion* [@Neal_2020, pp. 37], that is, variables that block all non-causal paths without introducing new ones [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014].
